Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-16 paper.qmd
Added MNISt code to a script

<!-- PaCMAP layout with author's suggested parameter choice-->
```{r}
#| label: read-mnist
## Import data
training_data_mnist <- read_rds("data/mnist/mnist_10_pcs_of_digit_1.rds")
names(training_data_mnist) <- paste0("x", 1:NCOL(training_data_mnist))

training_data_mnist <- training_data_mnist |>
  mutate(ID = 1:NROW(training_data_mnist))

data_mnist <- training_data_mnist |>
  select(-ID) |>
  mutate(type = "data")
```

@@

num_bins_x_mnist <- 30

algo_obj_mnist <- fit_highd_model(
  highd_data = training_data_mnist, 
  nldr_data = tsne_mnist2, 
  bin1 = num_bins_x_mnist, 
  q = 0.1, 
  benchmark_highdens = 1)

tsne_minst_scaled <- algo_obj_mnist$nldr_obj$scaled_nldr
tr_from_to_df_mnist <- algo_obj_mnist$trimesh_data
df_bin_centroids_mnist <- algo_obj_mnist$model_2d
df_bin_mnist <- algo_obj_mnist$model_highd

@@

tsne_minst_scaled <- tsne_minst_scaled |>
  mutate(cluster = if_else((emb1 <= 1) & (emb1 >= 0.78) & (emb2 <= 0.62) & (emb2 >= 0.39), "small_clust", "big_clust"))

@@

## Compute error
error_df <- augment(
  highd_data = training_data_mnist,
  model_2d = df_bin_centroids_mnist,
  model_highd = df_bin_mnist)

## To join embedding
error_df <- error_df |>
  bind_cols(tsne_minst_scaled |> 
              select(-ID))

## To transform the data
## ggplot(data = error_df, aes(x=row_wise_abs_error)) + geom_density() ## right skewed

error_df <- error_df |>
  mutate(sqrt_row_wise_tot_error = sqrt(row_wise_total_error)) |>
  mutate(sqrt_row_wise_tot_error = standardize(sqrt_row_wise_tot_error))

@@

<!--bin1 = 22, bin2 = 17, b = 374, non_empty = 140-->

```{r}
#| label: data-prep-mnist-model-proj

hexID_mnist <- df_bin_centroids_mnist |>
  #dplyr::filter(std_counts > 0) |>
  dplyr::pull(hexID)

df_b_mnist <- df_bin_mnist |>
  dplyr::filter(hexID %in% hexID_mnist) |>
  dplyr::mutate(type = "model") ## Data with summarized mean

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b_mnist <- df_b_mnist[match(hexID_mnist, df_b_mnist$hexID),]

df_b_mnist_temp <- df_b_mnist

df_b_mnist <- df_b_mnist |>
  dplyr::select(-hexID)

# Apply the scaling
df_model_data_mnist <- bind_rows(data_mnist, df_b_mnist)
scaled_mnist <- scale_data_manual(df_model_data_mnist, "type") |>
  as_tibble()

scaled_mnist_data <- scaled_mnist |>
  filter(type == "data") |>
  select(-type)

scaled_mnist_data_model <- scaled_mnist |>
  filter(type == "model") |>
  select(-type)
```

```{r}
#| label: langevitour-best-fit-mnist-proj-tsne
#| eval: false

data_mnist_cluster <- data_mnist |>
  select(-type) |>
  mutate(type = tsne_minst_scaled$cluster)

df_model_data_mnist_n <- bind_rows(df_b_mnist, data_mnist_cluster)

langevitour::langevitour(df_model_data_mnist_n[1:(length(df_model_data_mnist_n)-1)],
                         lineFrom = tr_from_to_df_mnist$from,
                         lineTo = tr_from_to_df_mnist$to,
                         group = factor(df_model_data_mnist_n$type,
                                        c("big_clust", "small_clust", "model")),
                         levelColors = c("#999999",'#ff7f00', "#000000"))
```

```{r}
#| label: link-plts-mnist
#| eval: false

df_exe_mnist <- comb_all_data_model(
  highd_data = training_data_mnist,
  nldr_data = tsne_minst_scaled_temp,
  model_highd = df_b_mnist_temp,
  model_2d = df_bin_centroids_dens_clust)

show_link_plots(point_data = df_exe_mnist,
                edge_data = tr_from_to_df_dens_clust)
```

@@

projection <- cbind(
  c(-0.04295,0.42361,-0.36702,0.13103,-0.19678,-0.29621,0.00471,0.09323,0.18822,-0.18484),
  c(-0.36644,0.05058,0.18200,0.21860,0.13583,-0.03511,0.35339,0.15821,-0.26945,-0.27926))

proj_obj1 <- get_projection(projection = projection, 
               proj_scale = 2.5, 
               highd_data = scaled_mnist_data, 
               model_highd = scaled_mnist_data_model, 
               trimesh_data = tr_from_to_df_mnist, 
               axis_param = list(limits = 1, 
                                 axis_scaled = 1.1, 
                                 axis_pos_x = -0.92, 
                                 axis_pos_y = -0.92, 
                                 threshold = 0.065))

proj_obj1[["cluster"]] <- tsne_minst_scaled$cluster

@@

## Second projection
projection <- cbind(
    c(-0.33305,0.20223,0.06303,0.11124,0.13299,-0.30093,0.34082,0.15581,-0.08864,0.33573),
    c(0.53615,0.14969,-0.09393,-0.20053,0.07992,-0.34791,0.13057,-0.06814,-0.16737,0.03716))

proj_obj2 <- get_projection(projection = projection, 
               proj_scale = 2, 
               highd_data = scaled_mnist_data, 
               model_highd = scaled_mnist_data_model, 
               trimesh_data = tr_from_to_df_mnist, 
               axis_param = list(limits = 1, 
                                 axis_scaled = 1.1, 
                                 axis_pos_x = -0.83, 
                                 axis_pos_y = -0.83, 
                                 threshold = 0.065))

proj_obj2[["cluster"]] <- tsne_minst_scaled$cluster

@@

```{r}
#| label: fig-clust-mnist
#| fig-pos: H
#| fig-height: 5
#| fig-width: 15
#| out-width: 100%
#| fig-cap: "The tSNE layout of the MNIST digit 1 data shows a big nonlinear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster in \\gD{} (a). The MNIST digit 1 data ($n =  7877$ and $p = 10$) has a nonlinear structure in $10\\text{-}D$. Two \\gD{} projections from a tour on $10\\text{-}D$ reveal that the closeness of the clusters in $10\\text{-}D$ and the twisted pattern of the model fit with tSNE ($a_1 = 0.041$, $b = 659/585 \\text{ } (30, \\text{ }48)$). Video of the langevitour animation is available at <https://youtu.be/sUcGd57Swdg>. The brushing feature in the linked plots helps in visualizing the closeness of the small cluster to the big cluster. Videos of the linked plots are available at <>."

@@

## Data with pixel values
mnist_data <- read_rds("data/mnist/mnist_digit_1.rds")

mnist_data <- mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28)

img_right_top <- c(868, 1465, 1787, 2006, 4314)

pixels_gathered_within <-  mnist_data |>
  filter(instance %in% img_right_top)

right_top_img <- plot_digit_img(
  pixels_gathered_within, palette = "Blues 2", title_text = "b")

img_middle <- c(130, 2091, 7795, 163, 77)

pixels_gathered_within <-  mnist_data |>
  filter(instance %in% img_middle)

middle_img <- plot_digit_img(
  pixels_gathered_within, palette = "Greens 3", title_text = "c")

img_right_bottom <- c(2035, 7708, 7710, 4134, 5006)

pixels_gathered_within <-  mnist_data |>
  filter(instance %in% img_right_bottom)

right_bottom_img <- plot_digit_img(
  pixels_gathered_within, palette = "Purples 2", title_text = "d")

img_error_outside <- c(999, 3650, 4916, 2964, 5622)

pixels_gathered_outside <-  mnist_data |>
  filter(instance %in% img_error_outside)

imge_error_sample_outside <- plot_digit_img(
  pixels_gathered_outside, palette = "Grays", title_text = "e")

##### inside
img_error_inside <- c(5173, 6981, 7228, 6673, 3179) #, 6673, 3179, 7363, 7650, 5241, 4636

pixels_gathered_inside <-  mnist_data |>
  filter(instance %in% img_error_inside)

imge_error_sample_inside <- plot_digit_img(
  pixels_gathered_inside, palette = "Oranges", title_text = "f")

>>>>

@@

@@

tsne_minst_scaled <- read_rds("data/mnist/mnist_tsne_minst_scaled.rds")

@@

## Compute error
error_df <- read_rds("data/mnist/mnist_error_df.rds")

@@

@@

proj_obj1 <- read_rds("data/mnist/mnist_proj_obj1.rds")

@@

## Second projection
proj_obj2 <- read_rds("data/mnist/mnist_proj_obj2.rds")

@@

```{r}
#| label: fig-clust-mnist
#| fig-pos: H
#| fig-height: 5
#| fig-width: 15
#| out-width: 100%
#| fig-cap: "The tSNE layout of the MNIST digit 1 data shows a big nonlinear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster in \\gD{} (a). The MNIST digit 1 data ($n =  7877$ and $p = 10$) has a nonlinear structure in $10\\text{-}D$. Two \\gD{} projections from a tour on $10\\text{-}D$ reveal that the closeness of the clusters in $10\\text{-}D$ and the twisted pattern of the model fit with tSNE ($a_1 = 0.041$, $b = 659/585 \\text{ } (30, \\text{ }48)$). The link to the video of the langevitour animation is available in @tbl-links. The brushing feature in the linked plots helps in visualizing the closeness of the small cluster to the big cluster. The link to the video of the linked plots is available in @tbl-links."

@@

## Data with pixel values

img_right_top <- read_rds("data/mnist/mnist_img_right_top.rds")

right_top_img <- plot_digit_img(
  img_right_top, palette = "Blues 2", title_text = "b")

img_middle <- read_rds("data/mnist/mnist_img_middle.rds")

middle_img <- plot_digit_img(
  img_middle, palette = "Greens 3", title_text = "c")

img_right_bottom <- read_rds("data/mnist/mnist_img_right_bottom.rds")

right_bottom_img <- plot_digit_img(
  img_right_bottom, palette = "Purples 2", title_text = "d")

img_error_outside <- read_rds("data/mnist/mnist_img_error_outside.rds")

imge_error_sample_outside <- plot_digit_img(
  img_error_outside, palette = "Grays", title_text = "e")

##### inside
img_error_inside <- read_rds("data/mnist/mnist_img_error_inside.rds")

imge_error_sample_inside <- plot_digit_img(
  img_error_inside, palette = "Oranges", title_text = "f")
